---
title: "MTBLS2 Processing and Analysis with xcms, CAMERA and export to MetaboLights"
author: "Steffen Neumann, Andrea Thum and Christoph Böttcher"
date: "`r Sys.Date()`"
bibliography:
  - ./MTBLS2.bib
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{MTBLS2 Processing and Analysis with xcms, CAMERA and export to MetaboLights}
output:
  html_document:
    theme: united	
    toc: true
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{MTBLS2}
-->

```{r LibraryPreload, message=FALSE}
library(Risa)
library(xcms)
````

## Introduction 

Indole-3-acetaldoxime (IAOx) represents an early intermediate of the biosynthesis of a variety of indolic secondary metabolites including the phytoanticipin indol-3-ylmethyl glucosinolate and the phytoalexin camalexin (3-thiazol-2'-yl-indole). Arabidopsis thaliana cyp79B2 cyp79B3 double knockout plants are completely impaired in the conversion of tryptophan to indole-3-acetaldoxime and do not accumulate IAOx-derived metabolites any longer. Consequently, comparative analysis of wild-type and cyp79B2 cyp79B3 plant lines has the potential to explore the complete range of IAOx-derived indolic secondary metabolites.


Since 2006, the Bioconductor package [xcms](http://bioconductor.org/packages/release/bioc/html/xcms.html)
[(Smith et al, 2006)](http://www.ncbi.nlm.nih.gov/pubmed/16448051) provides a rich set of
algorithms for mass spectrometry data processing. Typically, xcms will
create an xcmsSet object from several raw data files in an assay,
which are obtained from the samples in the study.  
Allowed raw data formats are netCDF, mzData, mzXML and mzML.

In this vignette, we demonstrate the processing of the MTBLS2 dataset, 
which was described in [Neumann 2012](http://www.springerlink.com/content/l148485q75010101).

## A few global settings

A few things might be worth to define at the beginning of an analysis

```{r settings} 
## How many CPU cores has your machine (or cluster) ?
nSlaves=4

# prefilter <- c(3,200)  ## standard
prefilter=c(6,750)      ## quick-run for debugging

```

## Raw data conversion

This can be done with the vendor tools, or the open source proteowizard converter. The preferred format should be mzML or mzData/mzXML. An overview of formats (and problems) is available at the [xcms online](https://xcmsonline.scripps.edu/docs/fileformats.html) help pages.

## R and ISAtab

An ISAtab archive will contain the metadata description in 
several tab-separated files. (One of) the assay files contains the column ``Raw Spectral Data File``
with the paths to the mass spectral raw data files in one of the above formats. 

```{r rISA, cache=TRUE}
setwd(find.package("mtbls2"))

ISAmtbls2 <- readISAtab()
a.filename <- ISAmtbls2["assay.filenames"][[1]]

````

## ISAtab, Risa and xcms

With the combination of [Risa](http://bioconductor.org/packages/release/bioc/html/Risa.html) and xcms, we can convert the MS raw data in an ISAtab archive into an xcmsSet:

```{r PeakPicking, cache=TRUE, warning=FALSE}
mtbls2Set <- processAssayXcmsSet(ISAmtbls2, a.filename,
                                 method="centWave", prefilter=prefilter, 
                                 snthr=25, ppm=25, 
                                 peakwidth=c(5,12),
                                 nSlaves=nSlaves)
````




```{r sessionInfo}
sessionInfo()
```

