---
title: "MTBLS2 Processing and Analysis with xcms, CAMERA and export to MetaboLights"
author: "Steffen Neumann, Andrea Thum and Christoph Boettcher"
date: "`r Sys.Date()`"
bibliography:
  - ./MTBLS2.bib
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{MTBLS2 Processing and Analysis with xcms, CAMERA and export to MetaboLights}
output:
  html_document:
    theme: united	
    toc: true
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{MTBLS2}
-->

```{r LibraryPreload, message=FALSE}
library(Risa)
library(xcms)
library(CAMERA)
library(pcaMethods)
library(Heatplus)
library(sp)
library(IPO)
````

## Introduction 

Indole-3-acetaldoxime (IAOx) represents an early intermediate of the biosynthesis of a variety of indolic secondary metabolites including the phytoanticipin indol-3-ylmethyl glucosinolate and the phytoalexin camalexin (3-thiazol-2'-yl-indole). Arabidopsis thaliana cyp79B2 cyp79B3 double knockout plants are completely impaired in the conversion of tryptophan to indole-3-acetaldoxime and do not accumulate IAOx-derived metabolites any longer. Consequently, comparative analysis of wild-type and cyp79B2 cyp79B3 plant lines has the potential to explore the complete range of IAOx-derived indolic secondary metabolites.


Since 2006, the Bioconductor package [xcms](http://bioconductor.org/packages/release/bioc/html/xcms.html)
[(Smith et al, 2006)](http://www.ncbi.nlm.nih.gov/pubmed/16448051) provides a rich set of
algorithms for mass spectrometry data processing. Typically, xcms will
create an xcmsSet object from several raw data files in an assay,
which are obtained from the samples in the study.  
Allowed raw data formats are netCDF, mzData, mzXML and mzML.

In this vignette, we demonstrate the processing of the MTBLS2 dataset, 
which was described in [Neumann 2012](http://www.springerlink.com/content/l148485q75010101).

## A few global settings

A few things might be worth to define at the beginning of an analysis

```{r settings} 
## How many CPU cores has your machine (or cluster) ?
nSlaves=4

# prefilter <- c(3,200)  ## standard
prefilter=c(6,750)      ## quick-run for debugging

```

## Raw data conversion

This can be done with the vendor tools, or the open source proteowizard converter. The preferred format should be mzML or mzData/mzXML. An overview of formats (and problems) is available at the [xcms online](https://xcmsonline.scripps.edu/docs/fileformats.html) help pages.

## R and ISAtab

An ISAtab archive will contain the metadata description in 
several tab-separated files. (One of) the assay files contains the column ``Raw Spectral Data File``
with the paths to the mass spectral raw data files in one of the above formats. 

```{r rISA, cache=TRUE}
setwd(find.package("mtbls2"))

ISAmtbls2 <- readISAtab()
a.filename <- ISAmtbls2["assay.filenames"][[1]]

````

## ISAtab, Risa and xcms

With the combination of [Risa](http://bioconductor.org/packages/release/bioc/html/Risa.html) and xcms, we can convert the MS raw data in an ISAtab archive into an xcmsSet:

```{r PeakPicking, cache=TRUE, warning=FALSE}
mtbls2Set <- processAssayXcmsSet(ISAmtbls2, a.filename,
                                 method="centWave", prefilter=prefilter, 
                                 snthr=25, ppm=25, 
                                 peakwidth=c(5,12),
                                 nSlaves=nSlaves)
````


The result is the same type of xcmsSet object:

```{r xcmsSet}
show(mtbls2Set)
``` 

Several options exist to
quantify the individual intensities. For each feature,
additional attributes are available, such as the minimum/maximum and
average retention time and m/z values. 

## Grouping and Retention time correction

In the following steps, we perform a grouping: because the UPLC system used here 
has very stable retention times, we just use the retention time correction step 
as quality control of the raw data. After that, 'fillPeaks()' will integrate 
the raw data for those features, which were not detected in some of the samples.

```{r retcor}
mtbls2Set <- group(mtbls2Set, minfrac=1, bw=3)
retcor(mtbls2Set, plottype="mdevden")

```

## QC on peaks picked

A first QC step is the visual inspection of intensities across the samples.
Alternatively to a boxplot, one could also create histograms/density plots.

```{r QCintensity}
boxplot(groupval(mtbls2Set, value="into")+1, 
        col=as.numeric(sampclass(mtbls2Set))+1, 
        log="y", las=2)
``` 

## How to optimize xcms Parameters ?

[Strategy for Optimizing LC-MS Data Processing in Metabolomics: A Design of Experiments Approach](http://pubs.acs.org/doi/abs/10.1021/ac301482k)


[IPO: A Tool for automated Optimization of XCMS Parameters](https://github.com/glibiseller/IPO)

```{r IPO}

mzdatapath <- system.file("mzData", package = "mtbls2")

## Use just one input file to reduce runtime. 
## Full set *can* be used manually. 

mzdatafiles <- list.files(mzdatapath, recursive = TRUE, full.names=TRUE)[3]

peakpickingParameters <- getDefaultXcmsSetStartingParams('centWave')

#setting levels for min_peakwidth to 10 and 20 (hence 15 is the center point)
peakpickingParameters$min_peakwidth <- c(5,10) 
peakpickingParameters$max_peakwidth <- c(15,25)

peakpickingParameters$ppm <- c(30,50)

resultPeakpicking <- optimizeXcmsSet(files=mzdatafiles, 
                          params=peakpickingParameters, 
                          nSlaves=nSlaves, subdir='rsmDirectory')
resultPeakpicking

```




## Data imputation

After grouping, peaks might be missing/not found in some samples.
`fillPekas()` will impute them, using the consensus mz and RT 
from the other samples. 

```{r fillPeaks, message=FALSE, warning=FALSE, results='hide' }
mtbls2Set <- fillPeaks(mtbls2Set, nSlaves=nSlaves)
```

The final xcmsSet represents a rectangular matrix of mass spectral features,
which were detected (or imputed) across the samples. The dimensionality is M * N,
where M denotes the number of samples in the assay, and N the number
of features grouped across the samples. 

## QC with heatmap and PCA

QC of mass accuracy and retention time consistency

```{r plotQC}
plotQC(mtbls2Set)
```


## QC with heatmap and PCA

In addition to the boxplot for QC, we can also check a hierarchical clustering 
and the PCA of the samples. 

```{r QCPCA, fig.show='hold'}
sdThresh <- 4.0 ## Filter low-standard deviation rows for plot
data <- log(groupval(mtbls2Set, value="into")+1)

pca.result <- pca(data, nPcs=3)
plotPcs(pca.result, type="loadings", 
        col=as.numeric(sampclass(mtbls2Set))+1)

## For readable heatmaps, 
data <- data[apply(data, MAR=1, FUN=sd) > sdThresh,]
heatmap_2(data, scale = "none", col=bpy.colors(100), legend=2)

```

## Annotated diffreport

```{r CAMERA, warning=FALSE, results='hide'}

an <- xsAnnotate(mtbls2Set,
                 sample=seq(1,length(sampnames(mtbls2Set))),
                 nSlaves=nSlaves)

an <- groupFWHM(an)
an <- findIsotopes(an)  # optional but recommended.
an <- groupCorr(an,
                graphMethod="lpc",
                calcIso = TRUE,
                calcCiS = TRUE,
                calcCaS = TRUE,
                cor_eic_th=0.5)

## Setup ruleSet
rs <- new("ruleSet")

rs@ionlistfile <- file.path(find.package("mtbls2"), "lists","ions.csv")
rs@neutraladditionfile <- file.path(find.package("mtbls2"), "lists","neutraladdition.csv")
rs@neutrallossfile <- file.path(find.package("mtbls2"), "lists","neutralloss.csv")

rs <- readLists(rs)
rs <- setDefaultParams(rs)
rs <- generateRules(rs)

an <- findAdducts(an,
                  rules=rs@rules,
                  polarity="positive")
  
```

## Diffreport
```{r diffreport}
dr <- diffreport(mtbls2Set, sortpval=FALSE, filebase="mtbls2diffreport", eicmax=20 )
cspl <- getPeaklist(an)

annotatedDiffreport <- cbind(dr, cspl)

```

## Combine diffreport and CAMERA spectra

```{r diffreportPspec}
interestingPspec <- tapply(seq(1, nrow(annotatedDiffreport)),
                               INDEX=annotatedDiffreport[,"pcgroup"],
                               FUN=function(x, a) {m <- median(annotatedDiffreport[x, "pvalue"]);
                                                   p <- max(annotatedDiffreport[x, "pcgroup"]);
                                                   as.numeric(c(pvalue=m,pcgroup=p))},
                               annotatedDiffreport)

interestingPspec <- do.call(rbind, interestingPspec)
colnames(interestingPspec) <- c("pvalue", "pcgroup") 

o <- order(interestingPspec[,"pvalue"])

pdf("interestingPspec.pdf")
dummy <- lapply(interestingPspec[o[1:40], "pcgroup"],
                function(x) {plotPsSpectrum(an, pspec=x, maxlabel=5)})
dev.off()

```


## Retention time outlier visualisation 

Sometimes, especially with larger sample sets, the standard `retcor()` will complain with "Too few peak groups, reverting to linear method" or "Not enough well behaved peak groups even for linear smoothing of retention times". There are several possible reasons why this can happen.

The underlying `group()` was probably not performed with a large enough bw param, e.g. UPLC parameters used for an HPLC setup. For large sample sets (way more than one hundred) you can increase the `retcor()` parameters `missing` and `extra`, which allows more peaks to be considered "well behaved". The `retcor(method="obiwarp")` method does not rely one such peak groups, and uses raw spectra correlation. It creates an xcmsSet with corrected retention times.

```{r retentionTimeOutlier }
    # Check RMSD of retention time deviation
     xs <- mtbls2Set
     xs <- retcor(xs)

     rmsd <- mapply(function(raw, corrected) sqrt(sum( (raw-corrected)**2) ),
                    xs@rt$raw, xs@rt$corrected)
    names(rmsd) <- sampnames(xs)
  plot(rmsd, xaxt="n", las=1)
  axis(1, at=1:length(rmsd), labels=names(rmsd), las=2)
```

In addition, it is even possible to cluster the samples w.r.t. their retention time profiles, which can discriminate between samples with the same RMSD in different areas of the gradient:


```{r retentionTimeOutlierClustering}
     minlength <- min(sapply(xs@rt$raw, length ))
     devs <- sapply(1:length(xs@rt$corrected),
                    function(x) { (xs@rt$raw[[x]]-xs@rt$corrected[[x]])[1:minlength] })
     colnames(devs) <- sampnames(xs)
     ddevs <- dist(t(devs))
     hdevs <- hclust(ddevs)
     plot(hdevs)
```


## R, ISAtab, xcms and CAMERA revisited

These attributes and the intensity matrix could already be exported to conform 
to the specification for the ``metabolite assignment file'' 
in the mzTab format used in MetaboLights. Currently, this functionality is not included 
in xcms. A prototype snippet is the following:

``` {r assembleMAF}

pl <- annotatedDiffreport 

charge <- sapply(an@isotopes, function(x) {
  ifelse( length(x) > 0, x$charge, NA) 
})
abundance <- groupval(an@xcmsSet, value="into")


##
## load ISA assay files
## 

a.samples <- ISAmtbls2["samples.per.assay.filename"][[ a.filename ]]

##
## These columns are defined by mzTab
##

maf.std.colnames <- c("identifier", "chemical_formula", "description",
"mass_to_charge", "fragmentation", "charge", "retention_time",
"taxid", "species", "database", "database_version", "reliability",
"uri", "search_engine", "search_engine_score", "modifications",
"smallmolecule_abundance_sub", "smallmolecule_abundance_stdev_sub",
"smallmolecule_abundance_std_error_sub")

##
## Plus the columns for the sample intensities
##
all.colnames <- c(maf.std.colnames, a.samples)

##
## Now assemble new maf
##

l <- nrow(pl)

maf <- data.frame(identifier = character(l), 
                  chemical_formula = character(l), 
                  description = character(l), 
                  mass_to_charge = pl$mz, 
                  fragmentation = character(l), 
                  charge = charge, 
                  retention_time = pl$rt, 
                  taxid = character(l), 
                  species = character(l), 
                  database = character(l), 
                  database_version = character(l), 
                  reliability = character(l), 
                  uri = character(l), 
                  search_engine = character(l), 
                  search_engine_score = character(l),
                  modifications = character(l), 
                  smallmolecule_abundance_sub = character(l),
                  smallmolecule_abundance_stdev_sub = character(l), 
                  smallmolecule_abundance_std_error_sub = character(l),
                  abundance, stringsAsFactors=FALSE)
```

```{r exportMAF}

##
## Make sure maf table is quoted properly, 
## and add to the ISAmtbls2 assay file.
## 
maf_character <- apply(maf, 2, as.character)

write.table(maf_character, 
            file="/tmp/a_mtbl2_metabolite profiling_mass spectrometry_maf.csv",
            row.names=FALSE, col.names=all.colnames, 
            quote=TRUE, sep="\t", na="\"\"")

ISAmtbls2 <- updateAssayMetadata(ISAmtbls2, a.filename,
             "Metabolite Assignment File",
             "a_mtbl2_metabolite profiling_mass spectrometry_maf.csv")

write.assay.file(ISAmtbls2, a.filename)

```


```{r sessionInfo}
sessionInfo()
```

